\begin{center}
  {\Large \bf Executive Summary}
\end{center}

More than 80 years after its astrophysical discovery, the fundamental nature of dark matter remains one of the foremost open questions in physics.
Over the last several decades, an extensive experimental program has sought to determine the cosmological origin, fundamental constituents, and interactions mechanisms of dark matter. 
While the existing experimental program has largely focused on weakly-interacting massive particles, there is strong theoretical motivation to explore a broader set of dark matter candidates.
As the high-energy physics program expands to ``search for dark matter along every feasible avenue'' \citep{P5:2014}, it is essential to keep in mind that the only direct, empirical measurement of dark matter properties comes from astrophysical and cosmological observations.

The Large Synoptic Survey Telescope (LSST), a major joint experimental effort between NSF and DOE, provides a unique and impressive platform to pursue the study of fundamental dark matter physics.
While LSST was originally envisioned as the ``Dark Matter Telescope'' \citep{Tyson:2001}, in recent years studies of fundamental physics with LSST have been more focused on dark energy.
In this document, we reaffirm LSST's ability to test well-motivated theoretical models of dark matter: \ie, self-interacting dark matter, warm dark matter, baryon-scattering dark matter, ultra-light/axion-like dark matter, and primordial black holes. 

LSST will use Milky Way satellite galaxies and stellar streams detect and study the smallest dark matter halos, thereby probing the minimum mass of ultra-light dark matter and thermal warm dark matter.
Precise measurements of the density and shapes of dark matter halos in dwarf galaxies and galaxy clusters will be sensitive to self-interacting dark matter with a cross section to probe hidden sector and dark photon models.
Microlensing measurement will directly probe primordial black holes and the compact object fraction of dark matter at the sub-percent level over a wide range of masses.
Precise measurements of stellar populations will be sensitive anomalous energy loss mechanisms and constrain couplings between photons and electrons to axion-like particles.
Unprecedented measurements of large scale structure will spatially resolve the influence of both dark matter and dark energy, enabling searches for correlations between the only empirically confirmed components of the dark sector.
In addition, complementarity between LSST, direct, and indirect dark matter experiments will help constrain dark matter-baryon scattering, dark matter self-annihilation, and dark matter decay.

The study of dark matter with LSST presents a small experimental program with a short timescale and low cost that is guaranteed to provide critical information about the fundamental nature of dark matter over the next decade.
LSST will produce high-impact science on fundamental dark matter physics by exploiting an existing US facility. 
The study of dark matter with LSST explores dark matter parameter space beyond the high-energy physics program's current sensitivity, while being highly complementary to other experimental searches. % BRN text.
This has been recognized during the Snowmass Cosmic Frontier planning process \citep[\eg,][]{1305.1605, 1310.8642, 1310.5662} and in a series of more recent Cosmic Visions reports \citep[\eg,][]{1604.07626,1802.07216}, including the ``New Ideas in Dark Matter 2017:\ Community Report'' \citep{1707.04591}.
It is worth remembering that astrophysical probes provide the only constraints on the minimum and maximum mass-scale of dark matter, and it is likely that astrophysical observations will continue to guide the  experimental particle physics program for years to come.

\clearpage

\begin{center}
  {\Large \bf Preface}
\end{center}

This white paper is the product of a large, global community of scientists who are united in support of probing the fundamental nature of dark matter with LSST.
Although LSST was originally proposed as the ``Dark Matter Telescope'' \citep{Tyson:2001}, none of the eight existing LSST Science Collaborations is specifically focused on exploring the microscopic identity of dark matter.
Rather, the study of dark matter is distributed across several disparate groups, making it difficult to combine results and build a cohesive physical picture of dark matter.
It was recognized that the current situation would severely hamper the progress of dark matter physics with LSST, and this effort was started to coordinate across disparate groups, to enlarge the dark matter community, and to strengthen connections between theory and experiment.

%This white paper arose
The concept for this white paper emerged from a series of meetings and regular telecons held in 2017--2018 around the topic of astrophysical probes of dark matter in the era of LSST.
Sessions were held at the LSST Project and Community Workshop in 2017 and 2018, multiple LSST Dark Energy Science Collaboration (DESC) meetings, and two dedicated multi-day workshops at the University of Pittsburgh and Lawrence Livermore National Laboratory.
Funding was provided by individual institutions and through an LSST Corporation (LSSTC) Enabling Science grant.
Through participation in the workshops, numerous telecons, sensitivity analyses, writing, editing, and reviewing, more than \CHECK{100} have directly contributed to this edition of the LSST dark matter white paper.

The current document is provide as an \emph{advanced draft}, and is currently under internal review within the LSST Dark Matter Group.

%with searches at collider, direct, and indirect particle physics experiments.

%This white paper summarizes the workshop “U.S. Cosmic Visions:  New Ideas in Dark Matter”  held  at  University  of  Maryland  from  March  23-25.   
%The  flagships  of  the  US  Dark Matter search program are the G2 experiments ADMX, LZ, and SuperCDMS, which will cover well-motivated axion and WIMP dark matter over a range of masses. 
%The workshop assumes that a complete exploration of this parameter space remains the highest priority of the dark matter community, and focuses instead on the science case for additional new small-scale  projects  in  dark  matter  science  that  complement  the  G2  program  (and  other ongoing projects worldwide).  
%It therefore concentrates on exploring distinct, well-motivated parameter space that will not be covered by the existing program;  on surveying ideas for such projects (i.e.  projects costing 10M or less); and on placing these ideas in a global context.  
%The workshop included over 100 presentations of new ideas, proposals and recent science and R&D results from the US and international scientific community.

% From the CMB-S4 science book
%This Science Book is the product of a large, global community of scientists who are united in support of proceeding with CMB-S4, which will make key advances in our understanding of the fundamental nature of space and time and the evolution of the Universe. The CMB-S4 concept was conceived during the 2013 Snowmass Cosmic Frontier planning exercise. Through the Snowmass process including two meetings and numerous telecons, the CMB experimental groups and the broader cosmology community came together to produce two influential CMB planning papers, endorsed by over 90 scientists, that outlined the science case as well as the CMB-S4 instrumental concept [1, 2]. It became clear that an enormous increase in the scale of ground-based CMB experiments would be needed to achieve the exciting scientific goals, necessitating a phase change in the ground-based CMB experimental program. To realize CMB-S4, a partnership of the university-based CMB groups, the broader cosmology community and the national laboratories would be needed.

%Based on the Snowmass papers and with additional information from the CMB experimental groups, the 2014 report of the Particle Physics Project Prioritization Panel (P5) included CMB-S4 in their recommended program. After the P5 report was released, the CMB community began a series of semi-annual workshops to advance CMB-S4. The first of these was a dedicated session at the Cosmology with the CMB and its Polarization workshop at the University of Minnesota January 14-16, 2015 attended by over 90 scientists. Discussions focused on the unique and vital role of the future ground-based CMB program and its synergy with a possible future satellite mission. It was decided that the community would draft a detailed CMB-S4 Science Book. The second and third workshops, Cosmology with CMB-S4 held at the University of Michigan September 21-22, 2015 with over 100 participants, and Cosmology with CMB-S4 held at the Lawrence Berkeley National Laboratory March 7-9, 2016 with over 160 participants, were dedicated to developing the Science Book. Working groups for each CMB-S4 science thrust were responsible for preparing and leading dedicated sessions at the workshop and for drafting the corresponding chapters of this book. In addition, a small writing group was responsible for integrating the Science Book. Through the workshops, numerous teleconferences, postings on the CMB-S4 wiki, contributions to the github Science Book repository, and feedback on drafts, over 200 scientists have contributed to this first edition of the CMB-S4 Science Book.
